---
title: "White Rice vs. Brown Rice"
output: pdf_document
header-includes:
  - \usepackage{booktabs}
---

```{r global_options, R.options=knitr::opts_chunk$set(warning=FALSE, message=FALSE)}
```

In this report, I'm using data from https://fdc.nal.usda.gov/index.html to persent a comparison of the nutritional 
value of brown rice vs. white rice. For information on this API look here. https://fdc.nal.usda.gov/api-guide.html


My API key is stored in .Renviron in my home directory using .Renviron as
data_gov_api_key

Here are two functions written to fetch the information.
I've learned from the [data documentation](https://www.ars.usda.gov/ARSUserFiles/80400530/pdf/fndds/2017_2018_FNDDS_Doc.pdf) that the food nutrient measures are per 100g of edible portion. Therefore, we need to scale the values by, a factor of portion weight / 100 in order to get he actual serving numbers for comparison.

Although the /v1/foods/search endpoint provides nutrient values, we need to fetch the /v1/foods endpoint to obtain the portion weights in orfer to normalize values based on portion size. So we have 

### Search Functions
```{r messages=F}
library(httr)

get.fdc <- function(path, ids, terms) {
  # append path
  fdc_url <- "https://api.nal.usda.gov"
  fdc_url <- modify_url(fdc_url, path = path)
  
  # build query string
  if (grepl("search$", path)) {
    qs <- paste(
      paste(c("query=description:", terms), collapse = " "),
      "requireAllWords=1",
      "dataType=Survey (FNDDS)",
      "pageSize=200",
      sep = "&")
    fdc_url <- modify_url(
      fdc_url,
      query = qs
      )
  } else {
    fdc_url <- modify_url(
      fdc_url,
      query = paste0("fdcIds=", paste(ids, collapse = ","))
      )
  }
  fdc_url <- URLencode(fdc_url)
  # print(fdc_url)

  r <- GET(
    fdc_url,
    add_headers('X-Api-Key' = Sys.getenv("data_gov_api_key"))
    )
  content(r)
}

```

### Retrieve Data
This code snippet will yield a data frame of four different kinds of rice: white, brown, yellow and wild.
```{r  message=FALSE, warning=FALSE}
library(tidyverse)

# First get the ids based on a descriptive search
rice1 <- get.fdc("fdc/v1/foods/search",
  terms = c(
    "rice", "cooked", "\"no added fat\"",
    "-\"cream of\"",
    "-\"brown and wild\"",
    "-\"white and wild\"")
  )
names(rice1$foods[[1]])
ids <- sapply(rice1$foods, function(x) x$fdcId)
rice_names <- sapply(rice1$foods,
                function(x) {
                  sub(
                    "^.*(brown|white|yellow|wild).*$",
                    "\\1 rice",
                    x$lowercaseDescription
                    )
                  }
                )
ids
rice_names

# then get the actual food details
rice2 <- get.fdc("fdc/v1/foods", ids = ids)

# examine the structure
str(rice2, max.level = 1)
names(rice2[[1]])

```
So now we can construct a usable data frame from the secondary results

```{r}
library(dplyr)
library(data.table)

portions <- sapply(rice2, function(x) x$foodPortions)
portionWeights <- sapply(portions[2,], function(x) x$gramWeight)

rice.df <- data.frame(
  id = ids,
  name = rice_names,
  portionMass = portionWeights 
)

rice.df


# portions <- as.data.frame(
#   sapply(
#     rice2.df,
#     function(x) {
#       unlist(x$foodPortions[[2]])
#       }[]
#     )
#   )
# 
# t_portions <- transpose(portions)
# colnames(t_portions) <- rownames(portions)
# rownames(t_portions) <- colnames(portions)
# t_portions2 <- t_portions %>%
#   mutate( servingScaleFactor = as.numeric(gramWeight) / 100 ) %>%
#   select(portionDescription,gramWeight, servingScaleFactor)
# t_portions2

#portions[row.names(portions) == c("gramWeight", "portionDescription"),]
#inputs

```


### Initial Inputs
```{r}
library(dplyr)
library(kableExtra)

white_rice_ingredients <-  foodItems[[1]]$inputFoods %>%
  transmute(
    ingredient=ingredientDescription,
    portion=portionDescription,
    grams=ingredientWeight
    ) 
brown_rice_ingredients <-  foodItems[[2]]$inputFoods %>%
  transmute(
    ingredient=ingredientDescription,
    portion=portionDescription,
    grams=ingredientWeight
    )
```

```{r echo=FALSE, results='asis'}
library(knitr)
library(xtable)

t1 <- white_rice_ingredients %>%
  kable(format = "latex", booktabs = TRUE) %>%
  column_spec(1, width = "11em")
t2 <- brown_rice_ingredients %>%
  kable(format = "latex", booktabs = TRUE)  %>%
  column_spec(1, width = "11em")

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.5\\linewidth}
      \\centering
        \\caption{White Rice Ingredients}",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.5\\linewidth}
      \\centering
        \\caption{Brown Rice Ingredients}",
        t2,
    "\\end{minipage} 
\\end{table}"
))

```

### Compare Nutrients
```{r echo=TRUE}
library(dplyr)
library(knitr)
library(xtable)

portions <- unlist(lapply(foodItems, function(x) {x$foodPortions$portionDescription[1]}))
names(portions) <- shortNames

white.nutrients.df <- white.df$foodNutrients[[1]] %>%
  transmute(nutrient=nutrientName,value,unit=unitName)

brown.nutrients.df <- brown.df$foodNutrients[[1]] %>%
  transmute(nutrient=nutrientName,value,unit=unitName)

combined.df <- full_join(
  white.nutrients.df, brown.nutrients.df, 
  by="nutrient", suffix=c(".white",".brown")
  ) %>%
  filter(
    unit.brown=="G" |
    unit.white == "G" |
    unit.brown == "KCAL" |
    unit.white == "KCAL"
  ) %>%
  filter(value.brown > 0, value.white > 0) %>%
  arrange(desc(value.white)) %>%
  rename(white=value.white, brown=value.brown) %>%
  select(nutrient, white, brown)

kable(combined.df, format = "latex", label="Table 3",
      caption="Nutrient List", booktabs = TRUE) 

```

### Calories
```{r}

compared.df <-data.frame(
  rbind(combined.df$white, combined.df$brown),
  row.names = shortNames)
names(compared.df) <- combined.df$nutrient

grams <- colSums(combined.df[2:6,2:3], na.rm=TRUE)
print(grams)

barplot(compared.df$Energy,  main = "Calories", names.arg=shortNames)
```

### Nutrients by Mass
```{r}
barplot(height=as.matrix(compared.df[,2:6]),names.arg=c("Water", "Carbohydrates", "Protein", "Fiber", "Lipids"), main="Nutrients", legend.text = shortNames, beside=TRUE, log = "y", args.legend = list(x = 'topright'))


```

